RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic Data in R

### 2013-12-24 inst/doc/manuscript.Rmd
writing writing

## Abstract

@@

@@

```{r}
phy <- get_trees(nex)
```
We can then leverage the rich suite of methods available from the R phylogenetics community (including coercion into other representations, such as `ouch` [@ouch] and `phylobase` [@phylobase] phylogeny formats). For instance, here we use the standard plotting function from the `ape` package to plot the resulting `ape::phylo` object returned by `get_trees`.

@@

@@

##### Character data

@@

### Writing NeXML phylogenies

```{r}
write.nexml(phy, "example.xml")
```

>>>>

---
title: RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic Data in R
author: 
  - name: Carl Boettiger
    affiliation: cstar
    email: cboettig@gmail.com
    footnote: Corresponding author
  - name: Hilmar Lapp
    affiliation: NESCent
  - name: Scott Chamberlain
    affiliation: SimonFraser
  - name: Rutger Vos
    affiliation: NCB
address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research, 
      Department of Applied Math and Statistics, 
      University of California, Mail Stop SOE-2,
      Santa Cruz, CA 95064, USA
  - code: NESCent
    address: | 
      National Evolutionary Synthesis Center, Durham, NC, USA
  - code: SimonFraser
    address: |
      Department of Biology, Simon Fraser University, Burnaby, Canada
  - code: NCB
    address: | 
      NCB Naturalis, Leiden, the Netherlands
abstract: | 
      1. NeXML promises to be the next generation of phylogenetic
      informatics data and metadata exchange format, offering substantial
      improvements in reliability, extensibility, and richness over
      earlier standards such as Nexus.

      2. In this paper we present the RNeXML package which provides
      a bridge between this rich and reliable data format and the
      extensive library of phylogenetic tools already available in R,
      with a particular emphasis on comparative methods.

      3. RNeXML allows R users to both read and write NeXML files. Yet
      RNeXML is much richer than a basic parser: here we illustrate
      how the package facilitates adding and manipulating semantic 
      annotations.

      4. An essential feature of NeXML is to provide both the flexibility
      to be continually extended to meet the emerging needs of the phylogenetics
      community while remaining backwards-compatible with software developed
      for earlier versions.  Here we illustrate how RNeXML makes it easy
      to extend NeXML in this way.

      5. RNeXML is perfectly suited for users to preserve and maintain their
      personal archives of phylogenetic and comparative trait data in consistent,
      query-able, and cross-platform format, or leverage the rapidly growing 
      public community archives of rich NeXML files.

---

Though R provides the most extensive and widely used collection of comparative phylogenetic methods tools, R users lack a convenient format to exchange and archive comparative methods data.

@@

- Most R users rely on the nexus inherits the limitations of nexus file format

@@

```{r}
phy <- get_trees(nex)
```

We can then leverage the rich suite of methods available from the R phylogenetics community (including coercion into other representations, such as `ouch` [@ouch] and `phylobase` [@phylobase] phylogeny formats). For instance, here we use the standard plotting function from the `ape` package to plot the resulting `ape::phylo` object returned by `get_trees`.

@@

#### Character data

Like nexus files, NeXML files can also contain character data.  While most researchers will be familiar with the use of character matrices in nexus files as a container for molecular sequence data used to infer the phylogenetic tree, we find this ideal for character data used in comparative methods popular in R. Methods to get extract character data work much like those for the phylogenetic data:

```{r }
nex <- nexml_read("comp_analysis.xml")
get_characters(nex)
```

Returns a `data.frame` with columns as characters and rows as taxa. Continuous and discrete characters are represented as seperate blocks in NeXML files, but will be combined as seperate columns of a single data frame by `get_characters()` if they correspond to the same taxa. The `get_characters_list()` function will consistently recieve a list of data frames corresponding to each character matrix block of the NeXML.

@@

Generating NeXML files from R is likewise straight forward.  Here we write a phylogeny from the `ape::phylo` format out to a NeXML file:

```{r}
data(bird.orders)
nexml_write(bird.orders, file = "birds.xml")
```

The first argument to `nexml_write` can be a `phylo` phylogeny, `multiPhylo` list of phylogenies, or a `nexml` object -- anything that can be coerced into the `nexml` class.  As we have seen, NeXML files can contain multiple phylogenetic trees, as well as character matrices.  We can add both trees  and character data using `nexml_write`,

```{r}
data(geospiza)
nexml_write(trees = geospiza$phy, characters = geospiza$dat, file="geospiza.xml")
```

NeXML files can also be built up bit by bit using the `add_characters`, `add_trees`, and `add_meta`.  Each of these functions can add data to an existing `nexml` object or create a new nexml object if none is provided.  Here we first create a `nexml` object containing the phylogeny data, and then add the character data to it:

```{r}
nex <- add_trees(geospiza$phy)
nex <- add_characters(geospiza$dat, nex)
```

The data need not share the same taxa.  We can append additional phylogenies and characters corresponding to different taxa:

```{r}
data(primates)
nex <- add_trees(primates$phy, nex)
nex <- add_characters(primates$dat, nex)
```

<!-- explain OTU blocks somewhere -->

### Metadata

@@

### Writing NeXML phylogenies
### Reading NeXML metadata
